---
title: "Lab_08"
author: "Kenya Sanchez"
format: pdf
editor: visual
---

# Lab 08 - April 22, 2026 

## Kenya Sanchez

## **Section 1: Overview**

> **radius:** mean of distances from nucleus center to points on the perimeter;

> **texture:** a measure of nucleus roughness taken from the standard deviation of gray-scale values;

> **perimeter:** total boundary length of the nucleus,

> **area:** total area of the nucleus;

> **smoothness:** local variation in radius lengths, i.e. how “bumpy” the edge is;

> **compactness:** measures how circular vs. irregular the shape is;

> **concavity:** how deeply indented, and

> **symmetry:** how symmetric the nucleus is.

## **Section 2: Exploratory data analysis**

### Getting organized

*Question 11 and 15 are optional extensions*
They can be skipped

### Preparing the data

Download and Import Data

```{r}
# Save your input data file into your Project directory
fna.data <- "WisconsinCancer.csv"

# Complete the following code to input the data and store as wisc.df
wisc.df <- read.csv(fna.data, row.names=1)
```

Examine your input data to ensure column names are set correctly

```{r}
# View the first 4 rows 
head(wisc.df, 4)
```

New Data Frame that Omits "Diagnosis" column

```{r}
# We can use -1 here to remove the first column
wisc.data <- wisc.df[,-1]
```

Generate a vector for the "Diagnosis" data

```{r}
# Create diagnosis vector for later 
diagnosis <- factor(wisc.df$diagnosis)
```

### Exploratory data analysis

Explore the data you created before (wisc.data and diagnosis) to answer the following questions:

**Q1: How many observations are in this dataset?**

> **Answer Q1:** Each row represents a different observation and a different patient. Each column represents a variable that describes the observation made. In this dataset there are a total of 569 observation.

```{r}
# #Observations = #Rows
dim(wisc.df)
nrow(wisc.df)
```

**Q2: How many of the observations have a malignant diagnosis?**

> **Answer Q2:** There are a total of 212 observations with a malignant diagnosis

```{r}
# Use table() function to reveal diagnosis of observations
table(diagnosis)

# Call out Malignant diagnosis ONLY
table(diagnosis)["M"]
```

**Q3: How many variables/features in the data are suffixed with \_mean?**

> **Answer Q3:** There are 10 variables/features suffized with "\_mean" in this data set.

```{r}
# How many variables / features present total 
length(wisc.df)
# Specifcy variables with suffic _mean
length(grep("_mean", colnames(wisc.df)))
```

## **Section 3: Principal Component Analysis**

### Performing PCA

Check Mean and Standard Deviation of features

```{r}
# Check column means and standard deviations
colMeans(wisc.data)

apply(wisc.data,2,sd)
```

Execute PRC

```{r}
# Perform PCA on wisc.data by completing the following code
wisc.pr <- prcomp(wisc.data, scale.=TRUE)
```

Summary of Results

```{r}
# Look at summary of results
summary(wisc.pr)
```

**Q4: From your results, what proportion of the original variance is captured by the first principal component (PC1)?**

> **Answer Q4:** The proprtion of original variance captured by the first principal component (PC1) is 44.2%. *PC1 capture 0.442 of the total variance in the data, 44.2% variance.*

```{r}
# Call our Importance table to find Proportion Variance of PC1
summary(wisc.pr)$importance[2,1]
```

**Q5: How many principal components (PCs) are required to describe at least 70% of the original variance in the data?**

> **Answer Q5:** In order to describe at least 70% of the original variance we would need the first 3 Principal Components.

```{r}
# Use sum() function to find the first components that describe at least 70%
# of original variance in data 

sum(summary(wisc.pr)$importance[3,]<=0.70) + 1
```

**Q6: How many principal components (PCs) are required to describe at least 90% of the original variance in the data?**

> **Answer Q6:** Nine principal componenets are needed in order to describe at least 90% of the orignal variance in the data.

```{r}
# Use sum() function to find the first components that describe at least 70%
# of original variance in data 

sum(summary(wisc.pr)$importance[3,]<=0.90) + 1
```

### Interpreting PCA results

Create a biplot of the wisc.pr using the biplot() function.

```{r}
biplot(wisc.pr)
```

**Q7: What stands out to you about this plot? Is it easy or difficult to understand? Why?**

> **Answer Q7:** The thing that stands out the most is that the row names are being ploted as characters. This biplot is very difficult to understand. Since rownames are used as the plotting character in the biplot shown above, we are unable to see a clear representation of a trend.

Generate Standard Scatter Plot

```{r}
# Call out ggplot2 from library 

# Scatter plot observations by Components 1 & 2

library(ggplot2)

ggplot(wisc.pr$x) +
  aes(x=PC1,y=PC2 , col=diagnosis) +
  geom_point()

```

**Q8: Generates a similar plot fir principal components 1 and 3. What do you notice about these plots?** 

> **Answer Q8:**  Shown below is the plot for PC1 and PC3 and when comparing these results to the plot above (PC1 and PC2) the separation between benign (B) and malignant (M) sampes is still visible. Overall, both plots help indicate that component 1 is important for capturing a separation between the begning (B) and the malignant (M) samples. In other words, PC1 helps fiveide out data into 2 clusters that separate the benign and malignant sample diagnoses. 

```{r}
# Repeat for components 1 and 3

ggplot(wisc.pr$x) +
  aes(PC1, PC3, col=diagnosis) +
  geom_point()


```


### Variance explained

Calculate Variance of components 

```{r}
# Calculate variance of each component

pr.var <- wisc.pr$sdev^2
head(pr.var)
```

Calculate the variance explained by: PC#/TotVar

```{r}
# Variance explained by each principal component: pve
pve <- pr.var / sum(pr.var)

# Plot variance explained for each principal component
plot(c(1,pve), xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

Alternative Histogram

```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Percent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

### Communicating PCA results

Loading Vector (wisc.pro$rotation) tells us which original measurements contribute most to each Principal Component 

**Q9: For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?** 

> **Answer Q9:** The component of the loading vector for the feature "concave.points_mean" is -0.2608538. There are about 30 features with contributions of a larger value than that of concave.points_mean. Show below are the results for the component loading vector of concave.points_mean as well as the features with larger contributions. 

```{r}
# Loading for concave.points_mean on PC1
wisc.pr$rotation["concave.points_mean", 1]

## Chech features with larger contributions
sort(abs(wisc.pr$rotation[,1]), decreasing = TRUE)
```

## **Section 4: Hierarchical clustering**

Scaling data ans assigning results 

```{r}
# Scale the wisc.data data using the "scale()" function
data.scaled <- scale(wisc.data)

```

Calculate Distances and assign results 

```{r}
data.dist <- dist(data.scaled)
```

Create Hierarchical Clustering 

```{r}
wisc.hclust <- hclust(data.dist, method="complete")
```

### Results of hierarchical clustering

Create a Cluster Dendrogram

**Q10: Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?**

> **Answer Q:10** The approximate height that results in 4 clusters is h=19. As shown in the R code and diagram below, past height 19 we see the cluster dendrogram section into 4 different clusters. 

```{r}
plot(wisc.hclust)
abline(h=19, col="red", lty=2)
```

### Selecting number of clusters

Evaluate clusters using  Cutree Function

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=4)
```

Table: compare Cluster Membership to Actual diagnosis 

```{r}
table(wisc.hclust.clusters, diagnosis)
```

**Q11: OPTIONAL: Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 6? How do you judge the quality of your result in each case?** 

> **Answer Q11:** *Skipped* 

### Using different methods

Different "*methods*" are used to combine points during hierarchical clustering. 

**Q12: Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.** 

> **Answer Q12:** My favorite methods are "complete" and "ward.D2" as these two show us the clearest models. My top favorite would be "ward.D2" because the clusters seem much more organized than any other method and the rownames (although many of them) appear to be more level. The bottom up orientation makes the graph look appealing. On the other hand, my least favorite was the average method, I understnad it dimishes the affect of outliers but it has a very large variance and makes the cluster dendrogram harder to understand. 

## **Section 5: Combining methods**

### Clustering on PCA results

Combining PCA and Hierarchical Clustering 
- First identify the minimum number of PCs required to describe *at least 90%* of the variability in the data. Then create an H-clustering model with linkeage method being "ward.D2". 


H-Cluster Dendrogram: 
```{r}
# How many PCs are required for 90% mark
sum(summary(wisc.pr)$importance[3,]<=0.90) + 1

# Combine PCA and H-cluster

wisc.pr.hclust <- hclust(dist(wisc.pr$x[,1:7]),
                  method = "ward.D2" ) 

plot(wisc.pr.hclust)
```
*Two main branches indicate two main clusters (B? & M?) 

```{r}
# Divide the Dendrogram: 2 main clusters
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```

```{r}
# Compare cluster to actual diagnosis 
table(grps, diagnosis)
```
This indicated that cluster 1 is predominantely malignant and cluster 2 is predominantely benign. 

```{r}
# Plot PCA results colored by 2 h-cluster  
ggplot(wisc.pr$x) +
  aes(PC1, PC2) +
  geom_point(col=grps)

```
Note: This plot looks very similar to that of the Malignant and Benign sample plot. 
Skipped 3D because will not work well with PDF. 

Copied example from lab to answer questions 13 and 14 
```{r}
## Use the distance along the first 7 PCs for clustering i.e. wisc.pr$x[, 1:7]
wisc.pr.hclust <- hclust(dist(wisc.pr$x[,1:7]), method="ward.D2")

wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=2)

table(wisc.pr.hclust.clusters, diagnosis)
```

**Q13: How well does the newly created hclust model with two clusters separate out the two “M” and “B” diagnoses?**

>**Answer Q13:** This newly created h-clust model with two clusters seaprates out the "M" and "B" diagnoses very accurately. The majority of the "Malignant" samples are found in cluster 1 and the majority of the "Benign" clusters are found in cluster 2. Although the separation is not perfect, it is still a very well model considering this was a "blind" analysis ( we did not provide the answers for B & M separation).

```{r}
# Compare to actual diagnoses
table(wisc.pr.hclust.clusters, diagnosis)
```


**Q14: How well do the hierarchical clustering models you created in the previous sections (i.e. without first doing PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.hclust.clusters and wisc.pr.hclust.clusters) with the vector containing the actual diagnoses.**

>**Answer Q14:** The hierarchical clustering models created in previous section (without PCA) are not as effective in terms of separating the diagnoses. We see this in the uneven and inaccurate distributions amongst the 4 clusters. Overall the Benign samples are clustered out very well but the malignant samples seem to be spread across multiple samples. This prevents an accurate separation between the two samples: M & B. (SHOWN BELOW)

```{r}
table(wisc.hclust.clusters, diagnosis)
```


## **Section 6: Sensitivity/Specificity**

**Q15: OPTIONAL: Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?** 

>**Answer Q15:** *Skipped*

## **Section 7: Prediction**

Use predict function to take PCA model from before and NEW cancer cell data and project that NEW data onto out PCA

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```
This tables reveals that in PC1 cluster 1 (which is predominantley malignant as stated before) has a much larger value than cluster 2. This means that cluster 1 will be found more to the right and cluster be will be towards the left on this plot. 


Plot New Data 

```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

**Q16: Which of these new patients should we prioritize for follow up based on your results?** 

>**Answer Q16:** Based on these results, we should prioritize the follow up for Patient 1. When these new patients were overlayed onto the existic PCA space, we see that patient 1 fell in cluster group 1 which has been identified as predominantly malignant. On the other hand, patient 2 falls in cluster 2 which is predominantly benign. Patient 1's location on the (predominantly) "Malignent" cluster indicates that they are at higher risk and that they require immediate medical attention; therefore, Patient 1 should be prioritized for a follow up consultaiton. 
